---
title: "Example ChIP-seq analysis"
author: "Pierre-Luc"
date: "3/10/2022"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(Rsubread)
  library(rtracklayer)
  library(Biostrings)
  library(Rfastp)
  library(epiwraps)
})
ah <- AnnotationHub()
```

We will perform the primary analysis of an example dataset, at each step demonstrating the use of both a very standard pipeline, and a full R-based equivalent. Pay attention to which chunks are R or bash code (see header)

# Downloading the raw reads

We download an example dataset which is a ChIP-seq experiment mapping the bindings
of the factor Myc across the Kc167 Drosophila cell line.
We download the data from ENCODE:

```{r, eval=FALSE}
dir.create("raw")
download.file("https://www.encodeproject.org/files/ENCFF014MOQ/@@download/ENCFF014MOQ.fastq.gz", dest="raw/Myc.fastq.gz")
download.file("https://www.encodeproject.org/files/ENCFF179TMO/@@download/ENCFF179TMO.fastq.gz", dest="raw/input.fastq.gz")
```

(Downloads have a timeout, so if your connection is slow or you're downloading a large file, you might need to change that, using for instance `options(timeout=600)`)

# Reads QC and trimming

## Using fastqc 

For this we will use shell commands:

```{bash engine.opts='-l', eval=FALSE}
mkdir -p raw_fastqc
fastqc -o raw_fastqc --threads 4 raw/*.fastq.gz
```

This will create an html report with QC plots for each file. We could optionally aggregate those using multiQC:

```{bash engine.opts='-l', eval=FALSE}
multiqc -o raw_fastqc raw_fastqc
```


### Trimming

```{bash engine.opts='-l', eval=FALSE}
mkdir -p trimmed
# this would be for a single sample:
trimmomatic SE -threads 4 -summary trimmed/Myc.stats -phred33 raw/Myc.fastq.gz trimmed/Myc.fastq.gz ILLUMINACLIP:TruSeq3-SE:2:15:4:4:true LEADING:20 TRAILING:20 SLIDINGWINDOW:4:15 MINLEN:25
trimmomatic SE -threads 4 -summary trimmed/input.stats -phred33 raw/input.fastq.gz trimmed/input.fastq.gz ILLUMINACLIP:TruSeq3-SE:2:15:4:4:true LEADING:20 TRAILING:20 SLIDINGWINDOW:4:15 MINLEN:25
```

## Using Rfastp

```{r}
dir.create("rfastp.trimmed")
qc <- lapply(c(Myc="raw/Myc.fastq.gz", input="raw/input.fastq.gz"), FUN=function(x){
  Rfastp::rfastp(x, thread=4, overrepresentationAnalysis=TRUE,
                 outputFastq=file.path("rfastp.trimmed/",gsub("\\.fastq\\.gz$","",basename(x))))
})
```

This will create an html report with before/after QC plots, as well as a trimmed
and filtered fastq file. Furthermore, the R output object (equivalent to the .json 
stats also saved by the program) can be used to generate the QC plots inside a
markdown, for example:

```{r}
Rfastp::curvePlot(qc$Myc, curve="content_curves")
```


# Alignment

## Using Rsubread

### Building a genome index for mapping

This we have to do only once for a genome, and can then re-use across projects

```{r, eval=FALSE}
# we get the genome sequence from AnnotationHub
genome <- ah[["AH49674"]]
# we create a new directory that will contain the genome index
dir.create("BDGP6_genome")
# we write the genome sequence in fasta format
export(import.2bit(genome), "BDGP6_genome/genome.fasta.gz", compress=TRUE)
# we build a Rsubread index
Rsubread::buildindex("BDGP6_genome/rsubread", reference="BDGP6_genome/genome.fasta.gz")
```

### Alignment

```{r}
dir.create("aligned")
align.stats <- Rsubread::align(index="BDGP6_genome/rsubread", type="dna",
                               readfile1=c("rfastp.trimmed/Myc_R1.fastq.gz", 
                                           "rfastp.trimmed/input_R1.fastq.gz"),
                               output_file=c("aligned/Myc.bam","aligned/input.bam"),
                               nthreads=6, sortReadsByCoordinates=TRUE)
align.stats
```


## Using bowtie2

### Building a bowtie2 index

```{bash engine.opts='-l'}
# bowtie2-build [options] path/to/genome.fasta path/to/outputIndex
bowtie2-build --threads 4 BDGP6_genome/genome.fasta.gz BDGP6_genome/bowtie2
```

### Alignment

```{bash engine.opts='-l'}
mkdir -p aligned
(bowtie2 -p 4 -x BDGP6_genome/bowtie2 -U trimmed/Myc.fastq.gz) 2> aligned/Myc.bowtie2 |\
  samtools view -bS - | samtools sort -@4 -m 2G - > aligned/Myc.bam
samtools index aligned/Myc.bam
(bowtie2 -p 4 -x BDGP6_genome/bowtie2 -U trimmed/input.fastq.gz) 2> aligned/input.bowtie2 |\
  samtools view -bS - | samtools sort -@4 -m 2G - > aligned/input.bam
samtools index aligned/input.bam
```

# Peak calling

## Using MACS2

```{bash engine.opts='-l'}
mkdir -p peaks
macs2 callpeak --outdir peaks -n Myc --gsize dm -t aligned/Myc.bam -c aligned/input.bam
# macs3 works in the same way
```

We can then import the peaks in R:

```{r}
peaks <- rtracklayer::import("peaks/Myc_peaks.narrowPeak")
source("peaks/Myc_model.r")
```


## Using R

This alternative is currently still very beta when it comes to using the input control, so here we will do without it. For the purpose of this course, this is offered as an alternative for windows users who can't install MACS2/3:

```{r}
peaks <- callPeaks("aligned/Myc.bam", fragLength=50L)
# if we want to save it as a bed file:
rtracklayer::export.bed(peaks, "peaks/peaks.bed")
```

