---
title: "DNA methylation and enrichment analysis"
author: "Emanuel / Pierre-Luc"
output:
  html_document: default
  pdf_document: default
---

```{r}
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(epiwraps)
  library(ggplot2)
  library(rGREAT)
  library(AnnotationHub)
  library(ensembldb)
  library(bsseq)
  library(BiocParallel)
  library(edgeR)
  library(DMRcate)
  library(rtracklayer)
  library(sechm)
  library(pheatmap)
  library(viridis)
  library(data.table)
})

set.seed(40)
```


## Load the data

Download:
```{r, download}
options(timeout = 6000)
download.file("https://ethz-ins.org/content/w11_practical.zip", "w11_practical.zip")
unzip("w10.practical.zip")
```

The .bigwig files have already been reduced to chromosome one and only have to be loaded here:
```{r, load bigwigs}
tracksGr <- list("ATAC"="./w11_practical/ATAC.rds",
                 "CTCF"="./w11_practical/CTCF.rds",
                 "H3K27ac"="./w11_practical/H3K27ac.rds",
                 "H3K4me3"="./w11_practical/H3K4me3.rds",
                 "DNAme"="./w11_practical/DNAm.rds")
tracksGr <- lapply(tracksGr, readRDS)
```


## Load the Annotation data

```{r, annotation}
ah <- AnnotationHub()
ensdb <- ah[["AH89211"]] # GRCm38
```

Obtaining the promoter coordinates of chromosome 1:
```{r, load promoters}
chr1 <-  GRanges(seqnames=Rle(c("1")), 
                          ranges = IRanges(1, end=195471971))

# For simplicity we assume here that the region +/- 150 of a tss constitutes a promoters
tssMargin <- 200
promoterRegions <- promoters(ensdb, upstream=tssMargin, downstream=tssMargin,
                             filter=GRangesFilter(chr1))
```

gene body coordinates: 
```{r, load gene bodies}
geneBodies <- genes(ensdb, columns=c("gene_seq_start", "gene_seq_end"),
                    filter=GRangesFilter(chr1))
```

## Enriched Heatmaps

### Promoters
```{r}
seqlevelsStyle(promoterRegions) <- "UCSC"
smTss <- signal2Matrix(tracksGr, promoterRegions[1:2000], 
                       extend=1000, w=20, 
                       type="scale", smooth=TRUE)

plotEnrichedHeatmaps(smTss[c(1,3:5)], axis_name = c("start","end"), colors = c("white","red"))

#plotEnrichedHeatmaps(smTss[1:3:4], axis_name = c("start", "end"), colors = c("white","red"), scale_title="signal")+
#plotEnrichedHeatmaps(smTss[5], axis_name = c("start", "end"), colors = c("white","red"), scale_title="DNAme-signal")
```

Clustering
```{r}
cl <- clusterSignalMatrices(smTss$DNAm, k=2)
table(cl)

mycolors <- c("1"="#E69F00", "2"="#56B4E9") # row_split=cl, mean_color=mycolors
plotEnrichedHeatmaps(smTss[c(1,3:5)], 
                     axis_name = c("start","end"), 
                     row_split=cl,
                     scale_title="signal",
                     mean_color=mycolors,
                     colors = c("white","red"))
#plotEnrichedHeatmaps(smTss[5], 
#                     axis_name = c("start","end"), 
#                     scale_title="DNAme-signal",
#                     row_split=cl,
#                     mean_color=mycolors,
#                     colors = c("white","red"))
```

For the colors see: [Colorblind Color Palette (Discrete) and Scales](https://jrnold.github.io/ggthemes/reference/colorblind.html)

### TF-Bindings

```{r}
plotEnrichedHeatmaps(smTss, 
                     axis_name = c("start","end"), 
                     row_split=cl,
                     scale_title="signal",
                     mean_color=mycolors,
                     colors = c("white","red"))
#plotEnrichedHeatmaps(smTss[5], 
#                     axis_name = c("start", "end"), 
#                     scale_title="DNAme-signal",
#                     row_split=cl,
#                     mean_color=mycolors,
#                     colors = c("white","red"))
```

## Differential Methylation Testing

### Bsseq object
The Bisulfite-sequenncing (BS-seq) data we are looking is from the [bsseqData](https://bioconductor.org/packages/release/data/experiment/html/bsseqData.html) package.
It contains colon cancer samples with 3 patients with each a colon cancer and normal colon sample. 
Here we only look at chromosome 22. 

```{r, bsseq}
bs <- readRDS("./data_practical/bs.rds")
rowRanges(bs)
pData(bs)
```

### Testing

Get annotations (hs):
```{r}
# genes
ensdb <- ah[["AH109336"]]

chr22 <-  GRanges(seqnames=Rle(c("22")), 
                  ranges = IRanges(1, end=195471971))
genesChr22 <- genes(ensdb, columns=c("gene_seq_start", "gene_seq_end", "gene_name"),
                    filter=GRangesFilter(chr22))
seqlevelsStyle(genesChr22) <- "UCSC"

# promoters
tssMargin <- 200
promotersChr22 <- promoters(ensdb, upstream=tssMargin, downstream=tssMargin,
                             filter=GRangesFilter(chr22), columns=c("gene_name"))
seqlevelsStyle(promotersChr22) <- "UCSC"
```

Retrieve metyhlation levels and visualize:
```{r, look at bsseq objects}
metPr <- bsseq::getMeth(bs, 
                        regions=promotersChr22[1:100], 
                        what="perRegion")
colnames(metPr) <- colnames(bs)
rownames(metPr) <- promotersChr22$gene_name[1:100]
metPr <- metPr[!is.na(rowSums(metPr)),]

library(viridis)
library(pheatmap)
annotationCol <- as.data.frame(pData(bs)[,c("Type","Pair")])
rownames(annotationCol) <- colnames(metPr)
pheatmap::pheatmap(metPr, 
                   cluster_rows=TRUE,
                   cluster_cols=FALSE,
                   annotation_col=annotationCol,
                   show_rownames = TRUE,
                   color=rocket(10))
```

Differential methylation testing: 
```{r, dm testing}
design <- model.matrix(~Type+Pair, data=pData(bs)) # design matrix

# adapt for methylation data
methdesign <- modelMatrixMeth(design)

seqAnnot <- sequencing.annotate(bs, methdesign, all.cov=TRUE, coef="Typenormal")
dmrcateRes <- dmrcate(seqAnnot, C=2, min.cpgs = 10, pcutoff=0.01)

dmrRanges <- extractRanges(dmrcateRes, genome="hg38")
saveRDS(dmrRanges, "./w11_practical/dmr.rds")

DMR.plot(dmrRanges, dmr=1,phen.col=c(rep("red", 3), rep("green",3)), CpGs=bs, genome="hg38")
dmrRangesGenes <- dmrRanges[!is.na(dmrRanges$overlapping.genes)]
```

Obtain the coordinates of the genes within DMRs. 
```{r, get the DM genes}
# Get the genes within Differentially methylated regions
topIdx <- order(dmrRangesGenes$min_smoothed_fdr)[1:10]
genesDmr <- unlist(tstrsplit(dmrRangesGenes[topIdx]$overlapping.genes, split=", "))
genesDmr <- genesDmr[!is.na(genesDmr)]
dmrGenes <- genesChr22[genesChr22$gene_name %in% genesDmr]
dmrGenes
```

