---
title: "Clustering and enrichment analysis"
author: "Pierre-Luc"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(epiwraps)
  library(ggplot2)
  library(rGREAT)
})
```

Today we're going to work on ChIP-seq of 2 TFs and 2 histone modifications in a mouse cancer line (all restricted to chr1).

## Download the data

```{r, eval=FALSE}
options(timeout = 6000)
download.file("https://ethz-ins.org/content/w10.practical.zip", "w10.practical.zip")
unzip("w10.practical.zip")
list.files()
```

## Prepare the regions and the tracks

```{r}
tracks <- list.files(pattern="bw$")
regions <- rtracklayer::import.bed("regions.bed")
```

## Plot

```{r}
ml <- signal2Matrix(tracks, regions, extend=2000)
set.seed(123)  # to ensure that it gives the same results everytime
nf <- getNormFactors(tracks, useSeqLevels="chr1")
ml2 <- rescaleSignalMatrices(ml, scaleFactors=nf)
plotEnrichedHeatmaps(ml)
```

## Clustering

```{r}
set.seed(123)  # to ensure that it gives the same results everytime
cl <- clusterSignalMatrices(ml, k=4)
table(cl)
head(cl)
length(cl)
length(regions)

regions$cluster <- cl
```

Plotting the clusters:

```{r}
plotEnrichedHeatmaps(ml, row_split=cl, trim=0.99, colors=c("white","darkred"))
```

```{r}
cl2 <- clusterSignalMatrices(ml, k=2:10)
ggplot(cl2$varExplained, aes(k, varExplained)) + geom_line()
```



Adding colors:

```{r}
mycolors <- c("1"="red", "2"="blue", "3"="darkgreen", "4"="black")
plotEnrichedHeatmaps(ml, row_split=cl, mean_color=mycolors)
```

Plotting just the averages:

```{r}
d <- meltSignals(ml, splitBy=cl)
ggplot(d, aes(position, mean, colour=sample)) + geom_line(size=1.2) + facet_wrap(~split)
```


Clustering using relative signal instead:

```{r}
cl <- clusterSignalMatrices(ml, k=4, scaleRows = TRUE)
d <- meltSignals(ml, splitBy=cl)
ggplot(d, aes(position, mean, colour=sample)) + geom_line() + facet_wrap(~split)
plotEnrichedHeatmaps(ml, row_split = cl, scale_rows = "global")
```


## Enrichment analysis

Find what's enriched in one cluster with respect to the others:

```{r}
# we first split the regions by cluster:
split_regions <- split(regions, cl)

# we send the data to the GREAT server to run an enrichment analysis
job <- submitGreatJob(gr=split_regions[["4"]], bg=regions, species="mm10")
res <- getEnrichmentTables(job)
names(res)
```

We plot the top Biological Processes:

```{r, fig.width=9, fig.height=6}
bp <- res$`GO Biological Process`
head(bp)

ggplot(head(bp,15), aes(Hyper_Fold_Enrichment, reorder(name, Hyper_Adjp_BH), size=Hyper_Foreground_Region_Hits, color=-log10(Hyper_Adjp_BH))) + geom_point() + scale_color_viridis_c()
```

