---
title: "Week 9 - Differential binding analysis"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(chromVAR)
  library(SummarizedExperiment)
  library(edgeR)
  library(epiwraps)
  library(ggplot2)
})
```


Here we'll work with only a subset of the data from the GDVS paper, namely the ChIPseq from just 2 mutant lines (deletions) and 2 controls.

## Downloading the data

```{r, eval=FALSE}
download.file("https://ethz-ins.org/content/yy1.chr9.zip", "yy1.chr9.zip", mode="wb")
unzip("yy1.chr9.zip")
```


## Getting a consensus set of windows

```{r}
peaks <- list.files("peaks",full=TRUE)
names(peaks) <- gsub("\\.narrowPeak","",basename(peaks))
peaks <- lapply(peaks, FUN=rtracklayer::import)

merged_peaks <- reduce(unlist(GRangesList(peaks)))
```

Note that a good alternative to get consensus peaks across samples is to merge the reads from all samples, and call peaks on that.

## Getting the counts per peak

```{r}
bams <- list.files("aligned/",pattern="bam$", full=TRUE)
se <- chromVAR::getCounts(bams, peaks=merged_peaks, paired=FALSE)
colnames(se) <- gsub("\\.bam","",basename(bams))
# we set the conditions:
se$condition <- c("CTRL","CTRL","DEL","DEL")
```

## Standard differential analysis with edgeR

```{r}
row.names(se) <- as.character(granges(se))
dds <- DGEList(assay(se), group=se$condition)
dds <- calcNormFactors(dds)
dds <- estimateDisp(dds)

et <- as.data.frame(topTags(exactTest(dds), Inf))

# # note that we could use the formula interface to specify the model:
# mm <- model.matrix(~condition, data=as.data.frame(colData(se)))
# dds <- estimateDisp(dds, mm)
# fit <- glmQLFit(dds, mm)
# res <- as.data.frame(topTags(glmQLFTest(fit)))
# # however, for simple 2-group comparisons in edgeR the exact test above is normally a tiny bit more accurate
```

Quick look at the results:

```{r}
head(et)
ggplot(et, aes(logFC, -log10(FDR))) + geom_point()
```

If you think back to the biology of the dataset we are analyzing, does anything look odd to you?

## Testing alternative normalization methods

### Normalization based on library size

```{r}
dds$samples$norm.factors <- 1
dds <- estimateDisp(dds)
et <- as.data.frame(topTags(exactTest(dds), Inf))
ggplot(et, aes(logFC, -log10(FDR))) + geom_point()
```

### Normalization based on common peaks

```{r}
nf.ma <- getNormFactors(bams, peaks=peaks, method="MAnorm")
dds$samples$norm.factors <- nf.ma
dds$samples$lib.size <- 1
dds <- estimateDisp(dds)
et <- as.data.frame(topTags(exactTest(dds), Inf))
ggplot(et, aes(logFC, -log10(FDR))) + geom_point()
```

### Background normalization

```{r}
set.seed(123)
nf.bg <- epiwraps::getNormFactors(bams, method="background")
dds$samples$norm.factors <- nf.bg
dds$samples$lib.size <- 1
dds <- estimateDisp(dds)
et <- as.data.frame(topTags(exactTest(dds), Inf))
head(et)
ggplot(et, aes(logFC, -log10(FDR))) + geom_point()
```


## Looking at the top differentially-bound peaks

```{r}
tracks <- list.files("tracks/", full=TRUE)
tracks
sm <- signal2Matrix(tracks, rowRanges(se))
nf <- getNormFactors(tracks)
nf.ma <- getNormFactors(tracks, peaks = peaks, method = "MAnorm")
plotEnrichedHeatmaps(sm)
plotEnrichedHeatmaps(rescaleSignalMatrices(sm, nf.ma))

# just the most significant regions
sm <- signal2Matrix(tracks, as(head(row.names(et),20),"GRanges"))
plotEnrichedHeatmaps(rescaleSignalMatrices(sm, nf))
```

